Leadership across cultures and contexts: an exploratory analysis

### 2020-04-26 leadership_across_cultures_contexts.Rmd
Spell check

Despite the universality of the phenomenon of leadership, the nature of leadership across populations and across social contexts is highly diverse and the emergence of leaders and the functions they serve are expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. In the absence of institutionalized organizational structure leaders often take on essential roles in group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group-wide projects, and administering sanctions to individuals who do not appropriately contribute to collective actions, comply with group policies, or seriously violate social norms. @glowacki_leadership_2015 draw on empirical, ethnographic, and theoretical models of leader emergence and suggest in small-scale societies institutionalized leadership roles are likely to initially emerge to resolve conflicts between families. @roscoe_social_2009 draws on the ethnography of highland New Guinea in describing the modular structure of human societies and suggests the functions of community leaders in small-scale societies are structured by the joint demands of signaling military fromidability to rival groups and facilitating within-group conflict resolution and cooperation. For "big-men" in these societies, in addition to military prowess, "talents for oratory, mediation, and organization were instrumental in reducing their internal conflict and enhancing the ability of their members to coordinate their actions as a group" [@roscoe_social_2009, p. 104].

@@

There are some broad commonalities in the qualities of community leaders across culturally and economically diverse populations. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. There has been increasing interdisciplinary discussion and consensus on the closely related topics of social status, power, authority, rank, prestige, and leadership [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014]. We draw on discussions by @blader_whats_2014 and @von_rueden_roots_2014 and coneptualize social status as a relative indicator of an individuals social value by the group and their access to contested resources within the group. The relationship between leadership (i.e., social influence) and status (i.e., social value) is expected to be colinear [@von_rueden_leadership_2015; @van_vugt_genes_2020], however, if there is causality between the two it is likely to be complex, context-specific, and driven by underlying correlates.

@@

<!-- BENEFITS: DECISIONS IN THEIR INTERST -->
The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the opportunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their close social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019]. 
<!-- BENEFITS: HIGH SOCIAL STATUS -->
Leadership positions, including formalized institutionalized roles and informal ephemeral roles, are also often if not universally associated with increased social status. The causal relationship between high social status and community or political leadership is complex and likely to be context-specific and bi-directional, but leadership roles through their association with increased social status are widely associated with numerous social, material, and reproductive benefits [@von_rueden_roots_2014; @von_rueden_multiple_2008; @grammer_symptoms_1996; @cheng_psychology_2014]. 
<!-- BENEFITS: SOCIAL BENEFITS, SOCIAL NETWORK -->
Leaders and high-status individuals are also more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Across many industries and organizations leadership positions are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also viewed as sexually attractive, particularly among women [@buss_strategies_2006; @stanik_who_2010].
<!-- NEED BETTER CITATIONS ABOVE -->
<!-- BENFITS: MATING -->
Across diverse populations and organizations, community leaders, political leaders, and organization leaders tend to have greater mating opportunities and therefore, given the aforementioned male-bias in top leadership positions, these leaders also tend to have greater reproductive success. In a phylogenetic multilevel metaanalysis of 288 reported statistical associations among nonindustrial societies, @von_rueden_mens_2016 found that most measures of reproductive success were positively associated with male social status, independent of subsistence type or status measure.

@@

The operationalized qualities and functions (See Table \@ref(tab:qualfuncvars)) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all `r nrow(all_data)` text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Table \@ref(tab:qualfuncvars). ZG then followed the same process of coding the entire sample of `r nrow(all_data)` text records for the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` operationalized variables, as well as a measure of group structure coded as: *residential subgroup*, which includes informal groups of co-residents, age-based groups, performance groups; *kin group*, which includes lineages, phraties, and other kin-based organizations; *economic group*, which includes subsistence groups, market groups, and other groups with primarily economic goals; *military group*, which includes all groups related to inter-group conflcit; *religious group*, which includes groups collected for spiritual or supernatural purpose; *political group (community)*, which includes political groups at the level fo the community, i.e., the political leader can potentially directly interact with most community members such as most village headmen, or *political group (supracommunity)*, which includes political groups that encompass multiple residential communities, such as complex chiefdoms, regional political leaders, and kings or state-level leaders. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository (a data frame named *leader_text2*).

@@

Coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we code if there is evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table \@ref(tab:costbenefitvars). There are multiple instances of conceptual overlap between many of our researcher-coded variables, in particular between various qualities and related costs or benefits. The operationalization of costs and benefits generaly capture a wider range of evidence than related qualities and focus more on evidence of transaction rather than description. We return to this point in the limitations section.

@@

To suppliment results from the aformentioend methods, we also performed textual analyses...OTHER METHODS. NEED TO DO.

@@

As Figure \@ref(fig:grouptype) reveals, the distribution of text records by group type varies across subsistence type. Across all subsistence types, the most frequent group type is political groups at the community level, with the exception of agriculturalitsts in which supracommunity political groups are the most frequent group type. Leadership beyond the community level is relatively rare outside of agricultural populations and is especially rare among hunter-gatherers. Leadership within residential groups is most frequently associated with hunter-gatherers (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and leadership within kin groups is most frequently associated with hunter-gatherers (`r hg_kin_pct`% of the total hunter-gatherers sample) and horticulturalists (`r hort_kin_pct`% of the total horticulturalists sample).

@@

*High status* was the most frequent leader quality in the ethnographic record, with `r statusPercTxt`% of text records and `r statusPercCult`% cultures containing evidence that leaders were of high social status. To explore factors that distinguished these records from those with no evidence for status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occured in each text record (see SI for details). We then fit an elasticnet logistic regression model (with lasso penalty) of high status (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provide evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1:4]`*, and *`r rev(highstatus_plot$data$Word)[5]`*. See Figure \@ref(fig:highstatusplot).

@@

Leadership in residential subgroups (e.g., hunting party, work group) was relatively common among hunter-gatherers but relatively rare among agriculturists. Leadership at the surpacommunity level, in contrast, was relatively rare among hunter-gatherers but quite common among agriculturists. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese, a group of populations of Negrito hunter-gatherers of the Andaman Islands by [@az02-001, p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section).

@@

The evolution of leadership and followership is somewhat perplexing. Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership (and by extension, group living) must also outweigh the costs for leadership to evolve and be maintained. We therefore coded all benefits and costs of both leadership and followership to better understand which benefits and costs are most commonly associated with leading or following. Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported. The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to recieve mating benefits, relative to leaders or other types benefits. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. The major difference between leaders and followers lay in leaders' opportunities for increased social status and mating opportunities, relative to followers. See Figure \@ref(fig:varsupportcb).

@@

```{r regionsubsistence, fig.width=11, fig.cap="Proportion of text records for different regions and subsistence types. Size of rectangles is proportional tothe number of text records."}
ggplot(all_data) + geom_mosaic(aes(x = product(subsistence, region), fill = subsistence)) + scale_fill_viridis(discrete = T) + theme_minimal(15)
```

@@

## Disentangleing the relationship between leadership and high status

These data present an oppurtunity to explore the relationship between high status and leadership, as represented in ethnographic texts. To do so, we draw on the term-document matrix of our ethnographic sample which was developed by computing the frequencies of all "informative" words in each text record after removing all punctuation and English stop words (uninformative words such as the, is, at), stemming the remaining words (reducing inflections such as plurals and past tenses to the root, or stem, words). In this data matrix the columns are all the unique word stems (terms), rows were the text records (documents), and the values in each cell were the frequencies of each word stem in each text record. The coding of *high status* was joined with this term-document matrix. This process produced a matrix with `r nrow(leader_dtm)` rows with one for each text-record and `r ncol(leader_dtm)` columns with a column for each unique term and a column for the researcher-coded *high status* value ($0$ or $1$).

@@

Figure \@ref(fig:highstatusplot) depicts the `r length(highstatus_plot$data$Word)` words with non-zero (mostly positive) coefficents. The words most predictive of evidence for high status leaders are *`r rev(highstatus_plot$data$Word)[1:4]`*, and *`r rev(highstatus_plot$data$Word)[5]`*.

We then investigated if the proportion of text records with evidence for high status differed among the different levels of social and political organization. There were modest differences. Somewhat counterintuitively, the proportion of text records indicating high status was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus).

@@

>>>>

Despite the universality of the phenomenon of leadership, the nature of leadership across populations and across social contexts is highly diverse and the emergence of leaders and the functions they serve are expected to be closely tied to group structure and group goals [@van_vugt_cognitive_2007]. Among many of the ethnographically described small-scale societies, which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and leaders emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. In the absence of institutionalized organizational structure leaders often take on essential roles in group coordination and collective action, including facilitating conflict resolution within or between autonomous households or kin groups, managing individual investments to group-wide projects, and administering sanctions to individuals who do not appropriately contribute to collective actions, comply with group policies, or seriously violate social norms. @glowacki_leadership_2015 draw on empirical, ethnographic, and theoretical models of leader emergence and suggest in small-scale societies institutionalized leadership roles are likely to initially emerge to resolve conflicts between families. @roscoe_social_2009 draws on the ethnography of highland New Guinea in describing the modular structure of human societies and suggests the functions of community leaders in small-scale societies are structured by the joint demands of signaling military formidability to rival groups and facilitating within-group conflict resolution and cooperation. For "big-men" in these societies, in addition to military prowess, "talents for oratory, mediation, and organization were instrumental in reducing their internal conflict and enhancing the ability of their members to coordinate their actions as a group" [@roscoe_social_2009, p. 104].

@@

There are some broad commonalities in the qualities of community leaders across culturally and economically diverse populations. When strong community leaders emerge, they are most typically high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have influence [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. There has been increasing interdisciplinary discussion and consensus on the closely related topics of social status, power, authority, rank, prestige, and leadership [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014]. We draw on discussions by @blader_whats_2014 and @von_rueden_roots_2014 and conceptualize social status as a relative indicator of an individuals social value by the group and their access to contested resources within the group. The relationship between leadership (i.e., social influence) and status (i.e., social value) is expected to be colinear [@von_rueden_leadership_2015; @van_vugt_genes_2020], however, if there is causality between the two it is likely to be complex, context-specific, and driven by underlying correlates.

@@

<!-- BENEFITS: DECISIONS IN THEIR INTERST -->
The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the opportunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their close social partners [@garfield_evolutionary_2019; @kantner_identifying_2010; @hagen_leadership_2019]. 
<!-- BENEFITS: HIGH SOCIAL STATUS -->
Leadership positions, including formalized institutionalized roles and informal ephemeral roles, are also often if not universally associated with increased social status. The causal relationship between high social status and community or political leadership is complex and likely to be context-specific and bi-directional, but leadership roles through their association with increased social status are widely associated with numerous social, material, and reproductive benefits [@von_rueden_roots_2014; @von_rueden_multiple_2008; @grammer_symptoms_1996; @cheng_psychology_2014]. 
<!-- BENEFITS: SOCIAL BENEFITS, SOCIAL NETWORK -->
Leaders and high-status individuals are also more likely to be the recipients of social support and material benefits in times of need [@gurven_its_2000; @von_rueden_roots_2014; @sugiyama_illness_2004]. Across many industries and organizations leadership positions are associated with increased financial compensation [@tsui_income_2017]. High social status, material wealth, intelligence, and leadership status are also viewed as sexually attractive, particularly among women [@buss_strategies_2006; @stanik_who_2010].
<!-- NEED BETTER CITATIONS ABOVE -->
<!-- BENFITS: MATING -->
Across diverse populations and organizations, community leaders, political leaders, and organization leaders tend to have greater mating opportunities and therefore, given the aforementioned male-bias in top leadership positions, these leaders also tend to have greater reproductive success. In a phylogenetic multilevel meta-analysis of 288 reported statistical associations among nonindustrial societies, @von_rueden_mens_2016 found that most measures of reproductive success were positively associated with male social status, independent of subsistence type or status measure.

@@

The operationalized qualities and functions (See Table \@ref(tab:qualfuncvars)) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all `r nrow(all_data)` text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Table \@ref(tab:qualfuncvars). ZG then followed the same process of coding the entire sample of `r nrow(all_data)` text records for the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` operationalized variables, as well as a measure of group structure coded as: *residential subgroup*, which includes informal groups of co-residents, age-based groups, performance groups; *kin group*, which includes lineages, phraties, and other kin-based organizations; *economic group*, which includes subsistence groups, market groups, and other groups with primarily economic goals; *military group*, which includes all groups related to inter-group conflict; *religious group*, which includes groups collected for spiritual or supernatural purpose; *political group (community)*, which includes political groups at the level of the community, i.e., the political leader can potentially directly interact with most community members such as most village headmen, or *political group (supracommunity)*, which includes political groups that encompass multiple residential communities, such as complex chiefdoms, regional political leaders, and kings or state-level leaders. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository (a data frame named *leader_text2*).

@@

Coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we code if there is evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table \@ref(tab:costbenefitvars). There are multiple instances of conceptual overlap between many of our researcher-coded variables, in particular between various qualities and related costs or benefits. The operationalization of costs and benefits generally capture a wider range of evidence than related qualities and focus more on evidence of transaction rather than description. We return to this point in the limitations section.

@@

To supplement results from the aforementioned methods, we also performed textual analyses...OTHER METHODS. NEED TO DO.

@@

As Figure \@ref(fig:grouptype) reveals, the distribution of text records by group type varies across subsistence type. Across all subsistence types, the most frequent group type is political groups at the community level, with the exception of agriculturalists in which supracommunity political groups are the most frequent group type. Leadership beyond the community level is relatively rare outside of agricultural populations and is especially rare among hunter-gatherers. Leadership within residential groups is most frequently associated with hunter-gatherers (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and leadership within kin groups is most frequently associated with hunter-gatherers (`r hg_kin_pct`% of the total hunter-gatherers sample) and horticulturalists (`r hort_kin_pct`% of the total horticulturalists sample).

@@

*High status* was the most frequent leader quality in the ethnographic record, with `r statusPercTxt`% of text records and `r statusPercCult`% cultures containing evidence that leaders were of high social status. To explore factors that distinguished these records from those with no evidence for status, we first created a document-term matrix (DTM) of all "informative" words in our corpus of texts and the frequency with which they occurred in each text record (see SI for details). We then fit an elasticnet logistic regression model (with lasso penalty) of high status (0/1) as a function of the frequencies of all `r ncol(leader_dtm)-1` words. Words that were strong positive predictors epitomized the semantic content of the text records which provide evidence of high status as a quality of leaders. These included *`r rev(highstatus_plot$data$Word)[1:4]`*, and *`r rev(highstatus_plot$data$Word)[5]`*. See Figure \@ref(fig:highstatusplot).

@@

Leadership in residential subgroups (e.g., hunting party, work group) was relatively common among hunter-gatherers but relatively rare among agriculturists. Leadership at the supracommunity level, in contrast, was relatively rare among hunter-gatherers but quite common among agriculturists. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese, a group of populations of Negrito hunter-gatherers of the Andaman Islands by [@az02-001, p. 45], who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders (see quote in Methods section).

@@

The evolution of leadership and followership is somewhat perplexing. Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership (and by extension, group living) must also outweigh the costs for leadership to evolve and be maintained. We therefore coded all benefits and costs of both leadership and followership to better understand which benefits and costs are most commonly associated with leading or following. Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported. The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other types benefits. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. The major difference between leaders and followers lay in leaders' opportunities for increased social status and mating opportunities, relative to followers. See Figure \@ref(fig:varsupportcb).

@@

```{r regionsubsistence, fig.width=11, fig.cap="Proportion of text records for different regions and subsistence types. Size of rectangles is proportional to the number of text records."}
ggplot(all_data) + geom_mosaic(aes(x = product(subsistence, region), fill = subsistence)) + scale_fill_viridis(discrete = T) + theme_minimal(15)
```

@@

## Disentangling the relationship between leadership and high status

These data present an opportunity to explore the relationship between high status and leadership, as represented in ethnographic texts. To do so, we draw on the term-document matrix of our ethnographic sample which was developed by computing the frequencies of all "informative" words in each text record after removing all punctuation and English stop words (uninformative words such as the, is, at), stemming the remaining words (reducing inflections such as plurals and past tenses to the root, or stem, words). In this data matrix the columns are all the unique word stems (terms), rows were the text records (documents), and the values in each cell were the frequencies of each word stem in each text record. The coding of *high status* was joined with this term-document matrix. This process produced a matrix with `r nrow(leader_dtm)` rows with one for each text-record and `r ncol(leader_dtm)` columns with a column for each unique term and a column for the researcher-coded *high status* value ($0$ or $1$).

@@

Figure \@ref(fig:highstatusplot) depicts the `r length(highstatus_plot$data$Word)` words with non-zero (mostly positive) coefficients. The words most predictive of evidence for high status leaders are *`r rev(highstatus_plot$data$Word)[1:4]`*, and *`r rev(highstatus_plot$data$Word)[5]`*.

We then investigated if the proportion of text records with evidence for high status differed among the different levels of social and political organization. There were modest differences. Somewhat counter-intuitively, the proportion of text records indicating high status was lowest at the highest level of political organization, perhaps because it would have been superfluous to note that kings and chiefs were high status. See Figure \@ref(fig:groupstatus).

@@

We constructed `r length(features)` "feature" variables from the higher-level clusters (Figure \@ref(fig:cluster)): `r paste(features, collapse = ', ')`. We then fit separate logistic mixed effects models of each feature as a function of subsistence mode, with random effects for author nested within culture. After adjusting the subsistence coefficient p-values for an FDR of 0.05, we computed and plotted the estimated marginal means of the remaining significant model(s). See Figure \@ref(fig:features).
